Reanalysis Step 2: Multiple Imputation

### 2025-08-30 backup/reanalysis-2-multiple_imputation_backup_20250829_173338.qmd
debug

# Additional packages for imputation
library(mice)
library(VIM)
library(naniar)

# ANALYSIS MODE: "test" or "final"
# test = 2 imputations, 2 iterations (fast)
# final = 30 imputations, 20 iterations (stability achieved ~20)
analysis_mode <- "final"  # Change to "final" for production run

@@

  # DR severity - use person_dr (created from No_DR/NPDR/PDR)
  "person_dr"  # Numeric 0/1/2
  # NOT dr_severity or person_dr_severity - derived from person_dr
)

df_mice <- df_for_imputation |>
  mutate(
    # Create a unique row identifier to preserve exact row matching
    row_id = row_number()
  ) |>
  dplyr::select(row_id, dplyr::all_of(imputation_vars)) |>
  mutate(
    # Convert person_dr to ordered factor for polr imputation method
    person_dr = factor(person_dr, levels = c(0, 1, 2), 
                      labels = c("No_DR", "NPDR", "PDR"),
                      ordered = TRUE),
    # Lock factor levels for categorical variables to ensure consistency
    cohort_id = factor(cohort_id),  # Ensure cohort_id is a factor
    gender_cat = factor(gender_cat),
    race_ethnic_cat = factor(race_ethnic_cat),
    insurance_cat = factor(insurance_cat)
  )
```

@@

# Temporal VA variables
custom_methods["baseline_VA_logMAR"] <- "pmm"  # Impute if missing
custom_methods["intermediate_VA_logMAR"] <- "pmm"  # IMPUTE intermediate values
custom_methods["outcome_VA_logMAR"] <- "pmm"  # Impute 2-year outcomes
custom_methods["cohort_id"] <- ""  # Never impute - data management indicator only

# Other continuous variables
custom_methods["CCI"] <- "pmm"  # Predictive mean matching for count data
custom_methods["DCSI"] <- "pmm"  # Predictive mean matching for count data
custom_methods["age"] <- "pmm"

@@

# Ensure exposure is used as predictor for outcomes
pred_matrix[, "ever_lapse_binary"] <- 1
pred_matrix["ever_lapse_binary", "ever_lapse_binary"] <- 0

@@

log_message("Starting MICE imputation process...")
imp_start <- Sys.time()

@@

# Check convergence
plot(mice_results)
```

## Imputation Diagnostics

**Note:** In the stripplots below, the leftmost column (labeled 1) shows the **original observed data** with missing values. Columns 2 through `r n_imputations + 1` show the imputed datasets.

```{r imputation-diagnostics}
# Check imputed values for key variables
# In mice stripplots: .imp = 1 is original data, .imp = 2+ are imputations

# CCI - compare observed vs imputed
if(sum(is.na(df_mice$CCI)) > 0) {
  # Density plot: Blue = observed data, Red = imputed values
  densityplot(mice_results, ~ CCI, 
              main = "Charlson Comorbidity Index (CCI)") |> print()

  # Stripplot: Column 1 = original data, Columns 2+ = imputed datasets
  stripplot(mice_results, CCI ~ .imp, pch = 20, cex = 1.2,
            xlab = "Dataset (1=Original, 2+=Imputations)",
            main = "CCI Distribution: Original vs Imputed") |> print()
}

# DCSI - compare observed vs imputed  
if(sum(is.na(df_mice$DCSI)) > 0) {
  densityplot(mice_results, ~ DCSI,
              main = "Diabetes Complications Severity Index (DCSI)") |> print()

  stripplot(mice_results, DCSI ~ .imp, pch = 20, cex = 1.2,
            xlab = "Dataset (1=Original, 2+=Imputations)",
            main = "DCSI Distribution: Original vs Imputed") |> print()
}

# Baseline VA - compare observed vs imputed
if(sum(is.na(df_mice$baseline_VA_logMAR)) > 0) {
  densityplot(mice_results, ~ baseline_VA_logMAR,
              main = "Baseline Visual Acuity (logMAR)") |> print()

  stripplot(mice_results, baseline_VA_logMAR ~ .imp, pch = 20, cex = 1.2,
            xlab = "Dataset (1=Original, 2+=Imputations)",
            main = "Baseline VA Distribution: Original vs Imputed") |> print()
}

# Intermediate VA - compare observed vs imputed
if(sum(is.na(df_mice$intermediate_VA_logMAR)) > 0) {
  densityplot(mice_results, ~ intermediate_VA_logMAR,
              main = "Intermediate Visual Acuity (logMAR)") |> print()

  stripplot(mice_results, intermediate_VA_logMAR ~ .imp, pch = 20, cex = 1.2,
            xlab = "Dataset (1=Original, 2+=Imputations)",
            main = "Intermediate VA Distribution: Original vs Imputed") |> print()
}

# Outcome VA - compare observed vs imputed
if(sum(is.na(df_mice$outcome_VA_logMAR)) > 0) {
  densityplot(mice_results, ~ outcome_VA_logMAR,
              main = "Outcome Visual Acuity at 2 Years (logMAR)") |> print()

  stripplot(mice_results, outcome_VA_logMAR ~ .imp, pch = 20, cex = 1.2,
            xlab = "Dataset (1=Original, 2+=Imputations)",
            main = "2-Year Outcome VA Distribution: Original vs Imputed") |> print()
}

# Person DR - compare observed vs imputed (categorical)
if(sum(is.na(df_mice$person_dr)) > 0) {
  # For categorical variables, check proportions across imputations
  # Extract each imputation and tabulate
  dr_props <- sapply(1:n_imputations, function(i) {
    imp_data <- complete(mice_results, i)
    prop.table(table(imp_data$person_dr))
  })

  # Create summary table
  dr_summary <- data.frame(
    Category = c("No_DR", "NPDR", "PDR"),
    Mean_Proportion = c(round(mean(dr_props[1,]), 3), 
                       round(mean(dr_props[2,]), 3),
                       round(mean(dr_props[3,]), 3)),
    SD = c(round(sd(dr_props[1,]), 3),
           round(sd(dr_props[2,]), 3), 
           round(sd(dr_props[3,]), 3))
  )
  kable(dr_summary, caption = "DR severity proportions across imputations")

  # Visual check if stripplot works with factors
  tryCatch({
    stripplot(mice_results, person_dr ~ .imp, pch = 20, cex = 1.2) |> print()
  }, error = function(e) {
    # Note: Stripplot not available for categorical person_dr
  })
}

# Check imputation quality for categorical variables
# Cataract surgery
if(sum(is.na(df_mice$catsurg_before_hitplus_cat)) > 0) {
  # Extract proportions for each imputation
  cat_props <- sapply(1:n_imputations, function(i) {
    imp_data <- complete(mice_results, i)
    prop.table(table(imp_data$catsurg_before_hitplus_cat))
  })

  # Create summary table for cataract surgery proportions
  if(is.matrix(cat_props)) {
    cat_summary <- data.frame(
      Level = rownames(cat_props),
      Mean_Proportion = apply(cat_props, 1, mean) |> round(3),
      SD = apply(cat_props, 1, sd) |> round(3)
    )
  } else {
    cat_summary <- data.frame(
      Level = "Cataract Surgery",
      Mean_Proportion = round(mean(cat_props), 3),
      SD = round(sd(cat_props), 3)
    )
  }
  kable(cat_summary, caption = "Cataract surgery proportions across imputations")
}
```

>>>>

# Additional packages for imputation
library(mice)
library(miceadds)  # For additional 2-level imputation methods
library(VIM)
library(naniar)

# ANALYSIS MODE: "test" or "final"
# test = 2 imputations, 2 iterations (fast)
# final = 30 imputations, 20 iterations (stability achieved ~20)
analysis_mode <- "test"  # Change to "final" for production run

@@

  # DR severity - use person_dr (created from No_DR/NPDR/PDR)
  "person_dr",  # Numeric 0/1/2
  # NOT dr_severity or person_dr_severity - derived from person_dr

  # Fellow eye predictors (if available)
  "fellow_baseline_VA",
  "fellow_intermediate_VA",
  "fellow_outcome_VA",
  # "fellow_lapsed",  # REMOVED: Collinear with ever_lapse_binary
  "fellow_dr",
  "n_eyes",
  # "has_fellow",  # REMOVED: Perfectly collinear with n_eyes (has_fellow = n_eyes == 2)
  "patient_worst_baseline",
  "patient_best_baseline"
  # "patient_ever_lapsed"  # REMOVED: Collinear with ever_lapse_binary
)

df_mice <- df_for_imputation |>
  mutate(
    # Create a unique row identifier to preserve exact row matching
    row_id = row_number()
  ) |>
  dplyr::select(row_id, dplyr::all_of(imputation_vars)) |>
  mutate(
    # Convert person_dr to ordered factor for polr imputation method
    person_dr = factor(person_dr, levels = c(0, 1, 2), 
                      labels = c("No_DR", "NPDR", "PDR"),
                      ordered = TRUE),
    # Lock factor levels for categorical variables to ensure consistency
    cohort_id = factor(cohort_id),  # Ensure cohort_id is a factor
    gender_cat = factor(gender_cat),
    race_ethnic_cat = factor(race_ethnic_cat),
    insurance_cat = factor(insurance_cat)
  ) |>
  mutate(
    # Clean factor levels to remove spaces and special characters for 2l.pmm compatibility
    # Spaces in factor levels cause formula parsing errors in lmer
    # Replace spaces and special characters with underscores
    across(where(is.factor), ~ {
      if (is.factor(.x)) {
        factor(.x, labels = gsub("[^[:alnum:]_]", "_", levels(.x)))
      } else {
        .x
      }
    })
  )
```

@@

# Debug: Check if any methods have invalid values
# Include all standard MICE methods plus 2-level methods from mice and miceadds
valid_methods <- c("", "pmm", "logreg", "polyreg", "polr", "norm", "norm.boot", "norm.nob", 
                   "norm.predict", "sample", "cart", "rf", 
                   # 2-level methods
                   "2l.norm", "2l.lmer", "2l.pan", "2lonly.mean", "2lonly.norm", "2lonly.pmm",
                   # miceadds methods
                   "2l.pmm", "2l.2stage.norm", "2l.2stage.pmm", "ml.lmer")
invalid_methods <- custom_methods[!custom_methods %in% valid_methods]
if (length(invalid_methods) > 0) {
  log_message(paste0("WARNING: Invalid imputation methods detected: ", paste(names(invalid_methods), "=", invalid_methods, collapse=", ")))
  # Reset any invalid methods to appropriate defaults
  for (var in names(invalid_methods)) {
    if (var %in% names(custom_methods)) {
      # Set to empty string to skip imputation for now
      custom_methods[var] <- ""
      log_message(paste0("Reset method for ", var, " to '' (no imputation)"))
    }
  }
}

# Ensure methods vector has proper names
if (is.null(names(custom_methods))) {
  log_message("ERROR: custom_methods has no names. Setting names from df_mice columns.")
  names(custom_methods) <- colnames(df_mice)
}

# Print diagnostic info
log_message(paste0("Number of variables in custom_methods: ", length(custom_methods)))
log_message(paste0("First few method assignments: ", 
                  paste(head(names(custom_methods)), "=", head(custom_methods), collapse=", ")))

# Temporal VA variables - use 2-level PMM for clustered data
# 2l.pmm from miceadds: Multilevel PMM using linear mixed models
# This properly accounts for correlation between eyes from same patient
custom_methods["baseline_VA_logMAR"] <- "2l.pmm"  # Multilevel PMM for eyes within patients
custom_methods["intermediate_VA_logMAR"] <- "2l.pmm"  # Multilevel PMM for eyes within patients  
custom_methods["outcome_VA_logMAR"] <- "2l.pmm"  # Multilevel PMM for eyes within patients
custom_methods["cohort_id"] <- ""  # Never impute - data management indicator only

# Other continuous variables - use 2l.pmm for clustered data
custom_methods["CCI"] <- "2l.pmm"  # Multilevel PMM
custom_methods["DCSI"] <- "2l.pmm"  # Multilevel PMM
custom_methods["age"] <- "pmm"  # Regular PMM (patient-level, doesn't vary by eye)

@@

# For 2l.pmm variables, use a selective set of predictors to avoid formula length issues
# Identify which variables will use 2l.pmm
vars_2l_pmm <- c("baseline_VA_logMAR", "intermediate_VA_logMAR", "outcome_VA_logMAR", "CCI", "DCSI")

# Define core predictors for 2l.pmm models (keep it focused)
core_predictors_2l <- c(
  "ever_lapse_binary",  # Exposure
  "age",                # Key demographic
  "gender_cat",         # Gender
  "baseline_VA_logMAR", # Baseline VA (for temporal)
  "intermediate_VA_logMAR", # Intermediate (for outcome)
  "person_dr",          # DR severity
  "anti_VEGF",          # Key treatment
  "CCI", "DCSI",       # Comorbidities
  "fellow_baseline_VA", # Fellow eye for correlation
  "fellow_outcome_VA"   # Fellow eye outcome
)

# For 2l.pmm variables, limit predictors to core set
log_message("Configuring predictor matrix for 2l.pmm variables with limited predictors")
for (var in vars_2l_pmm) {
  if (var %in% rownames(pred_matrix)) {
    # First set all to 0
    pred_matrix[var, ] <- 0
    # Then selectively enable core predictors
    n_predictors <- 0
    for (pred in core_predictors_2l) {
      if (pred %in% colnames(pred_matrix) && pred != var) {
        pred_matrix[var, pred] <- 1
        n_predictors <- n_predictors + 1
      }
    }
    log_message(paste0("  ", var, ": ", n_predictors, " predictors"))
  }
}

# Ensure exposure is used as predictor for all outcomes
pred_matrix[, "ever_lapse_binary"] <- 1
pred_matrix["ever_lapse_binary", "ever_lapse_binary"] <- 0

# DR severity predicts visual acuity at all timepoints
if ("person_dr" %in% colnames(pred_matrix)) {
  pred_matrix["baseline_VA_logMAR", "person_dr"] <- 1
  pred_matrix["intermediate_VA_logMAR", "person_dr"] <- 1
  pred_matrix["outcome_VA_logMAR", "person_dr"] <- 1
}

# FELLOW EYE STRUCTURE
# Fellow eye outcomes strongly predict index eye outcomes (correlation typically 0.7-0.9)
# NOTE: MICE predictor matrix only uses 0/1, not weights. 1 = used as predictor, 0 = not used
if ("fellow_outcome_VA" %in% colnames(pred_matrix)) {
  # Fellow eye visual acuity predicts same timepoint
  pred_matrix["outcome_VA_logMAR", "fellow_outcome_VA"] <- 1
  pred_matrix["baseline_VA_logMAR", "fellow_baseline_VA"] <- 1  
  pred_matrix["intermediate_VA_logMAR", "fellow_intermediate_VA"] <- 1

  # Fellow baseline also helps predict later timepoints
  pred_matrix["outcome_VA_logMAR", "fellow_baseline_VA"] <- 1
  pred_matrix["intermediate_VA_logMAR", "fellow_baseline_VA"] <- 1

  # DR severity correlation between eyes
  if ("person_dr" %in% rownames(pred_matrix)) {
    pred_matrix["person_dr", "fellow_dr"] <- 1
  }

  # Don't let fellow eye variables be predicted by anything
  # Only set for variables that actually exist in the matrix
  fellow_vars_in_matrix <- c("fellow_outcome_VA", "fellow_baseline_VA", "fellow_intermediate_VA", "fellow_dr")
  fellow_vars_in_matrix <- fellow_vars_in_matrix[fellow_vars_in_matrix %in% rownames(pred_matrix)]
  if (length(fellow_vars_in_matrix) > 0) {
    pred_matrix[fellow_vars_in_matrix, ] <- 0
  }

  # Don't let fellow eye variables be their own predictors
  if ("fellow_outcome_VA" %in% rownames(pred_matrix)) {
    pred_matrix["fellow_outcome_VA", "fellow_outcome_VA"] <- 0
  }
  if ("fellow_baseline_VA" %in% rownames(pred_matrix)) {
    pred_matrix["fellow_baseline_VA", "fellow_baseline_VA"] <- 0
  }
  if ("fellow_intermediate_VA" %in% rownames(pred_matrix)) {
    pred_matrix["fellow_intermediate_VA", "fellow_intermediate_VA"] <- 0
  }
}

# Patient-level variables can predict eye-level outcomes (but not strongly)
if ("patient_worst_baseline" %in% colnames(pred_matrix)) {
  pred_matrix[, "patient_worst_baseline"] <- 1
  pred_matrix[, "patient_best_baseline"] <- 1
  # pred_matrix[, "patient_ever_lapsed"] <- 1  # REMOVED - variable not in imputation

  # But don't predict these variables
  pred_matrix["patient_worst_baseline", ] <- 0
  pred_matrix["patient_best_baseline", ] <- 0
  # pred_matrix["patient_ever_lapsed", ] <- 0  # REMOVED - variable not in imputation
}

# n_eyes is just an indicator, not a predictor
if ("n_eyes" %in% colnames(pred_matrix)) {
  pred_matrix[, "n_eyes"] <- 0
  pred_matrix["n_eyes", ] <- 0
  # has_fellow removed from imputation - was collinear with n_eyes
}

@@

log_message("Starting MICE imputation process...")

# Final check on methods before running MICE
log_message("Final method check before imputation:")
methods_with_values <- custom_methods[custom_methods != ""]
log_message(paste0("Variables to be imputed: ", paste(names(methods_with_values), collapse=", ")))
log_message(paste0("Methods to be used: ", paste(unique(methods_with_values), collapse=", ")))

# Expected warnings about collinearity:
# 1. e_mrn_deidentified: Constant (patient ID) - expected, not used as predictor
# 2. outcome_VA_logMAR & fellow_outcome_VA: Identical missingness patterns (both NA for excluded cohort)
#    This is EXPECTED and HARMLESS - they measure different eyes but are missing for same structural reason
#    MICE will drop fellow_outcome_VA from the imputation model for outcome_VA_logMAR to avoid
#    singularity, but it can still be used to predict other variables. This does NOT affect results.
# 
# Variables removed to avoid true collinearity: has_fellow, patient_ever_lapsed, fellow_lapsed

imp_start <- Sys.time()

@@

# Save convergence plot for later diagnostics
pdf(file.path(reanalysis_data_dir, "mice_convergence.pdf"))
plot(mice_results)
dev.off()

log_message("Convergence plot saved to mice_convergence.pdf")
```
